\doctitle

### 2023-04-24 msr3.tex
Incorporate all Sasha and Jim's edits from the rebuttal

Variable selection is a fundamental problem in all regression settings. In linear regression, the LASSO method~\citep{tibshirani1996regression} and related extensions have been widely used. % for this purpose. 
However, variable selection for LMEs is complicated by the nonlinear structure and relative sparsity of the within-group data. 
While standard methods and software are available for linear regression (see e.g. \texttt{glmnet}~\cite{glmnet}), there are few open source libraries for variable selection 
for LMEs. 
Many covariates selection algorithms for LMEs have been proposed over the last 20 years (see the survey \cite{Buscemi2019Survey}), but comparison of 
these strategies and practical application remains difficult. %practical application of these strategies is difficult. 
%Most solutions share the goal of improving selection quality but vary significantly in implementation details. 
Approaches vary by choice of likelihood (e.g. marginal, restricted, or h- likelihood),  
regularizer~(e.g. $\ell_1$~\citep{Krishna2008} or SCAD~\cite{ibrahim2011fixed}), and information criteria \citep{Vaida2005,Ibrahim2011}. 
%and 
%sometimes  estimate effects in stages \citep{Krishna2008}. 
Implementations vary as well, typically using regularizer-specific local quadratic approximations to apply
solution methods for smooth problems (Newton-Raphson, EM, sequential least squares) to fit the original nonsmooth model. 
All of these decisions make it  difficult to compare and evaluate performance of available 
variable selection strategies and to determine which method is best suited for a given task. 
This challenge  is exacerbated by the absence of standardized datasets %for different experimental designs
and open source libraries for each method. 
Our main practical goal to fill this gap by developing a unified methodological framework that
accommodates a wide variety of variable selection strategies based on a set
of easily implementable regularizers, and implemented in an open source library that makes 
it easy to use and to compare different methods.

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/summary_picture.pdf}
    \caption{Selection of fixed and random effects for LME likelihoods $\mathcal{L}$ using 
    `regularization-agnostic' framework and its SR3 extension using four regularizers. 
    %standard and SR3-relaxed linear mixed-effect likelihoods with popular sparsity-promoting regularizers for the task of simultaneous 
    %The original likelihood $\LL(\beta, \gamma)$ (Eq. \ref{eq:lmm_objective}) is relaxed with SR3 approach to obtain $u(\beta, \gamma)$ (Eq. \ref{eq:value_function_definition}). Both setups are then solved using the same Proximal Gradient Descent (PGD) algorithm. 
    SR3 relaxation accelerates algorithmic converge  (middle panel), and gives better robustness and improved performance on synthetic problems across regularizers (right panel)}.
    \label{fig:summary}
\end{figure}

In this work we develop a regularization-agnostic covariate selection strategy that (1) is fast and simple to implement, (2) provides robust models, and (3) is flexible enough to support most regularizers currently used in variable selection across different domains.  The baseline approach uses the proximal gradient descent (PGD) method, which has been studied by the optimization community for over 40 years, but has not been widely used in LME covariate selection. We provide proximal operators for commonly used regularizers and show how to apply the PGD method to the nonconvex LME setting. In particular we apply the PGD method to four regularizers, 
including the $\ell_0$ regularizer, which does not admit local quadratic approximations and has not been used before for LME variable selection strategies.

We also develop a new meta-approach that can improve the performance of LME selection methods for any regularizer. Specifically, we extend sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting. In linear regression, SR3 accelerates and improves the performance of  regularization strategies by introducing auxiliary variables that decouple the accuracy and sparsity requirements for the model coefficients.
We develop a conceptual and algorithmic approach necessary to extend the SR3 concept to LME.  This development is necessary because the LME problem is nonlinear, nonconvex, and includes constraints on variance parameters.  We show that the new approach yields superior results in terms of specificity and sensitivity of feature selection, and is also computationally efficient.

\edit{Add a paragraph summarizing results from the theory paper}

All new methods are implemented in an open-source library called \texttt{pysr3}, which fills a gap for python mixed-models selection tools in \texttt{Python} (\cite{Buscemi2019Survey}, Table 3). Our algorithms are  1-2 orders of magnitude faster than available LASSO-based libraries for  mixed effects selection in \texttt{R},  see Table \ref{table:glmmlasso}. \texttt{pysr3} enables a standardized comparison of different methods in the LME setting, and makes both the PGD framework and its SR3 extension available to practitioners working with LME models.

\edit{Extend on the package summary using JOSS paper}

%\subsection{Notation}
%\begin{enumerate}
%\item sets like $\N$, $\B$, $\bS$, $\SS$.
%\item nonsmooth stuff, subdifferential and normal cones
%\end{enumerate}

\section{Linear Mixed-Effects Models: Notation and Fundamentals}
Mixed-effect models describe the relationship between an outcome variable and its predictors when the observations are grouped, for example in studies or clusters.  To set the notation, consider $m$ groups of observations indexed by $i$, with sizes $n_i$, and the total number of observations equal to $n = n_1 + n_2 + \dots + n_m$. For each group, we have design matrices for fixed features $X_i \in \R^{n_i \times p}$,  and matrices of random features $Z_i \in \R^{n_i \times q}$, along with vectors of outcomes $Y_i \in \R^{n_i}$. 
%Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general. 
Let 
{$X = [X_1^T, X_2^T, \dots, X_m^T]^T$ and $Z = [Z_1^T, Z_2^T, \dots, Z_m^T]^T$.} 
Following~\cite{Patterson1971, Pinheiro2000}, we define a Linear Mixed-Effects (LME) model as
\eq{
	\label{eq:lme_setup}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \bS_{+}^{q} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \bS_{++}^{n_i}
}
 where $\beta \in \R^p$ is a vector of fixed (mean) covariates, 
 $u_i \in \R^{q}$ are unobservable random effects assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$, and $\bS_{+}^{\nu}$ 
 and $\bS_{++}^{\nu}$ are the sets of
 real symmetric $\nu\times \nu$ positive semi-definite
 and positive definite matrices, respectively. 
Matrices $Z_i$ encode a wide variety of models, including
random intercepts ($Z_i$ are columns of 1's that add $u_i$ to all datapoints from the $i$th study)
and random slopes ($Z_i$ also scale $u_i$ according to the magnitude of a covarite), see e.g.~\cite{pinheiro2006mixed}. 
%In these typical examples, the columns of $Z_i$ are subsets of the columns
%of $X_i$.  
In our study, we assume that the observation error covariance matrices 
$\Lambda_i$ are given and that the random effects covariance matrix 
is an unknown diagonal matrix, i.e., $\Gamma = \Diag{\gamma}, \ \gamma \in \R^s_+$.
% either
% assumed to have a simple parametric form, such as $\Lambda_i=\sig_i I$
% with $\sig_i$ to be determiined, or given given  
 %; we focus on the latter case below.

Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
recasts~\eqref{eq:lme_setup} as a correlated noise model:
 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i.
}
For brevity, we refer to $\Omega_i(\Gamma)$ as just $\Omega_i$.
%without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
The reformulation \eqref{eq:lmm_correlated_noise_setup}
yields the following marginalized  negative log-likelihood function of a linear mixed-effects model~\citep{Patterson1971}:
\eq{
	\label{eq:lmm_objective}
	\LL_{ML}(\beta, \Gamma)  :=
%	 \textabove{$\Delta$}{=}
	 \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\ln{\det{\Omega_i}}.
}
Maximum likelihood estimates for $\beta$ and $\Gam$ are obtained by
solving the optimization problem
\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma)  \quad \mbox{s.t.} \quad \ \Gamma \in \bS_{+}^{q}.
}

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation 
%\eq{
%	
\[
\label{eq:beta_formula}
	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
\] %}
By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
%{In the meta-analysis setting under study, 
%In our study, we assume that $\Gam$ is a diagonal matrix of the form
%$\Gamma = \Diag{\gamma}, \ \gamma \in \R^s$ with the 
%$\Lambda_i \in \R_{++}^{n_i \times n_i}$ fixed and given 
%%the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
%so that our problem takes the form}
\eq{
	\label{eq:lme_diagonal_setup}
	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
	:=\LL_{ML}(\beta, \Diag{\gamma}) 
%	\\
%	s.t. & \ \gamma \geq 0\ .
	}
%Under the assumption $\Lambda_i \in \R_{++}^{n_i \times n_i}$ in~\eqref{eq:lme_setup}, 
In this setting, when an
entry $\gamma_j$ takes the  value $0$ the corresponding coordinates of all random effects $u_{ij}$ are identically $0$ for all $i$.

@@

This paper focuses the case where $\Gamma$ is diagonal, 
  (often referred to as \textit{the diagonal setup}) 
 and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely used in epidemiological studies~\cite{murray2020global}. 
While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.

@@

Variable (feature) selection models seeks to select or rank the most important predictors in a dataset in order to get a parsimonious model at a minimal cost to prediction quality. 
%The selection process often occurs simultaneously with fitting the model, and a predictor is `selected' if its respective coefficient in the model is non-zero. 
If the desired number of coefficients $k$ is given, then the feature selection problem can be formulated as the minimization of a loss function $f(\theta)$ (e.g. the negative log-likelihood) subject to a zero-norm constraint:
\eq{
	\label{eq:general_feature_selection_zero_norm}
	&\min_{\theta}  f(\theta) \quad \mbox{ s.t. }  \quad \|\theta\|_0 \leq k 
	}
where $\|\theta\|_0$ denotes the number of nonzero entries in $\theta$, see panel (c) of Figure~\ref{fig:regularizers}.

@@

 \begin{figure}[ht]
    \centering
    \begin{tabular}{ccc}
        \includegraphics[width=0.3\textwidth]{figures/l1_regularizer.pdf}
&
           \includegraphics[width=0.3\textwidth]{figures/lh_regularizer.pdf}
    &
    \includegraphics[width=0.3\textwidth]{figures/l0_regularizer.pdf}\\
  %      \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
              (a) $\ell_1$  & (b) $\ell_{1/2}$ & (c) $\ell_0$
    \end{tabular}
    \caption{Common convex and non-convex regularizers used for feature selection.}
    \label{fig:regularizers}
\end{figure}

@@

 Feature selection for LMEs is more difficult than for linear regression models. 
 In linear regression the observations are independent, whereas in mixed-effects setup they are generally correlated. 
 In addition, LMEs have both mean effect variables  $\beta$ as well as random variance variables 
 $\Gamma$. % where the groups can have different relative importance. 
% Finally, selecting random features implies that $\Gamma$ will have zero columns and rows thereby forcing the optimization methods to the boundary of the feasible parameter set and making both the theoretical analysis and numerical computations more challenging.  
The shrinkage operator approach for linear regression 
\citep{Tibshirani1996} was first adapted 
to the problem of feature selection for the fixed effects in mixed-effect models by \cite{Lan2006}. 
The removal of a random effect from the model requires the elimination of an entire row and column from $\Gamma$. 
To make the problem more tractable, \cite{Chen2003} reparamtrized $\Gamma$ through a modified Cholesky decomposition $\Gamma(D,L): = DLL^TD$, where $D$ is a diagonal matrix and $L$ is a lower-triangular matrix with ones on the main diagonal, and focused on selecting elements of $D$. 
% In this case, one can select variance parameters by selecting the elements of the diagonal matrix $D$ while treating the non-diagonal entries of $L$ as free parameters. 
Based on this idea, \cite{Krishna2008} extended the Adaptive LASSO regularizer (\cite{Lan2006, Xu2015}) to mixed-effects setting using the objective
   \[
        \LL(\beta, \Gamma(D,L)) + \lambda\left(\sum_{i = 1}^p 
        \left|\frac{\beta_i}{\hat\beta_i}\right| + \sum_{j = 1}^q \frac{D_{ii}}{\hat{D_{ii}}} \right),
   \]
 where $\hat\beta$ and $\hat{D}$ are the solution of a non-penalized maximum likelihood problem and $\lambda$ is a tuning parameter for the weighted regularizer
 and is called the regularization parameter. 
 %This strategy has a well known issue: non-penalized estimates $\hat\beta$ and $\hat{D}$ may be inaccessible because the numerical algorithm may fail to converge, especially when the underlying solution is sparse. 
 \cite{Ibrahim2011} use a similar approach, penalizing  non-zero elements $\Gamma_{ij}$ directly. 
 Other methods that use Adaptive LASSO for simultaneous selection 
 of fixed and random effects are \cite{Lin2013, fan2014robust, pan2018simultaneous}. Adaptive LASSO is available to practitioners via \texttt{R} packages \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} (\cite{groll2014variable}) and \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}(\cite{schelldorfer2011estimation}).

>>>>

Variable selection is a fundamental problem in all regression settings. In linear regression, the LASSO method~\citep{tibshirani1996regression} and related extensions have been widely used. % for this purpose. 
However, variable selection for LMEs is complicated by the nonlinear structure and relative sparsity of the within-group data. 
While standard methods and software are available for linear regression (see e.g. \texttt{glmnet}~\cite{glmnet}), there are few open source libraries for variable selection 
for LMEs. 
Many covariates selection algorithms for LMEs have been proposed over the last 20 years (see the survey \cite{Buscemi2019Survey}), but comparison of 
these strategies and practical application remains difficult. %practical application of these strategies is difficult. 
%Most solutions share the goal of improving selection quality but vary significantly in implementation details. 
Approaches vary by choice of likelihood (e.g. marginal, restricted, or h- likelihood),  
regularizer~(e.g. $\ell_1$~\citep{Krishna2008} or SCAD~\cite{ibrahim2011fixed}), and information criteria \citep{Vaida2005,Ibrahim2011}. 
%and 
%sometimes  estimate effects in stages \citep{Krishna2008}. 
Implementations vary as well, typically using regularizer-specific local quadratic approximations to apply
solution methods for smooth problems (Newton-Raphson, EM, sequential least squares) to fit the original nonsmooth model. 
All of these decisions make it  difficult to compare and evaluate performance of available 
variable selection strategies and to determine which method is best suited for a given task. 
This challenge  is exacerbated by the absence of standardized datasets 
%for different experimental designs
and open source libraries for each method. 
Our main practical goal to fill this gap by developing a unified methodological framework that
accommodates a wide variety of variable selection strategies based on a set
of easily implementable regularizers, and 
made available in an open source library, \texttt{pysr3}\footnote{https://github.com/aksholokhov/pysr3} that is 
easy to use and to compare different methods. All experiments in the paper can be reproduced 
using \texttt{pysr3} and code in the reproducibility guide\footnote{https://github.com/aksholokhov/msr3-paper}.

\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth]{figures/summary_picture.pdf}
    \caption{Selection of fixed and random effects for LME likelihoods $\mathcal{L}$ using 
    `regularization-agnostic' framework and its SR3 extension using four regularizers. 
    %standard and SR3-relaxed linear mixed-effect likelihoods with popular sparsity-promoting regularizers for the task of simultaneous 
    %The original likelihood $\LL(\beta, \gamma)$ (Eq. \ref{eq:lmm_objective}) is relaxed with SR3 approach to obtain $u(\beta, \gamma)$ (Eq. \ref{eq:value_function_definition}). Both setups are then solved using the same Proximal Gradient Descent (PGD) algorithm. 
    SR3 relaxation accelerates algorithmic converge  (middle panel), and gives better robustness and improved performance on synthetic problems across regularizers (right panel)}.
    \label{fig:summary}
\end{figure}

In this work we introduce a regularization-agnostic covariate selection strategy that (1) is fast and simple to implement, (2) provides robust models, and (3) is flexible enough to support most regularizers currently used in variable selection across different domains.  The baseline approach uses the proximal gradient descent (PGD) method, which has been studied by the optimization community for over 40 years, but has not been widely used in LME covariate selection. 
%We provide proximal operators for commonly used regularizers and show how to apply the PGD method to the nonconvex LME setting. In particular we apply the PGD method to four regularizers, 
%including the $\ell_0$ regularizer, which does not admit local quadratic approximations and has not been used before for LME variable selection strategies. 
In our initial numerical experiments, using a naive PGD approach
indicated that, at best, the method yields only a marginal improvement over the
equally unsatisfactory alternative methods   
in accurately determining the correct variables in our variable selection test problems.
We conjecture that the weakness of the naive PGD is a result of the 
first-order likelihood approximation used in PGD.
To overcome this problem, we propose an alternative likelihood approximation 
with the goal of 
incorporating global variational properties of the likelihood. For this purpose,  
we extend the sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting. 
This idea is supported
by the success of SR3 in the context of linear regression where it
accelerates and improves the performance of regularization strategies.
This extension and its mathematical foundations in \cite{Theory1} constitute
the major innovations of this work.
Here we introduce the modeling framework and its relaxation, discuss the 
resulting algorithms and their implementation details, and  validate the method on both
simulated and real data sets, while the 
%highly technical 
mathematical 
foundations are presented in \cite{Theory1}.
The SR3 framework introduces auxiliary variables $x$ and $w$ to 
decouple the likelihood $\LL$ and sparsity regularizer $R$ which are tied
together by adding a multiple of the norm squared difference 
$\frac{\eta}{2}\norm{x-w}^2$ to the
objective. 
%We then smooth the positive semi-definite constraint in the 
%likelihood with a log barrier term yielding the smoothed likelihood $\mathcal{L}_\mu$
%where $\mu$ is the smoothing parameter.
Then, fixing the variables $w$ dedicated to the nonsmooth regularizer $R$, 
the smooth function $\LL(x)+\frac{\eta}{2}\norm{x-w}^2$ is
globally optimized over the variables $x$ to obtain an optimal value function
$u_\eta(w)$. We then show that $u_\eta(w)$ is smooth. This opens
the door to the application of the PGD algorithm to minimizing $u_\eta+R$ 
where now $u_\eta$ contains global variational information
on the likelihood function $\LL$. The main obstacle in the
application of this approach is the evaluation of $u_\eta$ and its gradient.
%%%
In Section \ref{sec:MSR3} we present a method for overcoming this difficulty 
using variable metric techniques and interior point technology.
%%%%%%
%To overcome this difficulty we turn to interior point technology. This choice requires 
%smoothing the positive semi-definite constraint on the covariance matrices by
%adding log-barrier term to the likelihood to obtain the smooth function
%$\LL_{\mu,\eta}$ where $\mu$ is the log-barrier relaxation parameter, and
%an associated optimal value function $u_\emu$ whose value and gradients 
%can be quickly and accurately computed.
%%by applying variable metric techniques along with
%%well established properties of the interior point (IP) algorithm to obtain 
%%sufficiently accurate values for
%%both $u_\eta$ and its gradient. 
%Using this underlying framework, we develop two related algorithms,
%MSR3 and MR3-fast where MSR3-fast employs a kind of continuation approach 
%that interlaces approximate IP optimality conditions for the evaluation of 
%$u_\emu$ and its gradient with the prox computation in the PGD algorithm.
%%using a {\it central path} proximity condition from the IP literature.  
%The resulting algorithms are then applied to four regularizers, 
%including the $\ell_0$ regularizer 
%which does not admit local quadratic approximations and 
%has not been used before for LME variable selection strategies.
%The numerical results show that both MSR3 and MR3-fast
%yield superior results over the naive PGD method in terms of specificity and sensitivity of feature selection
%with a slight edge to MR3-fast. However, MR3-fast delivers a spectacular improvement
%in speed over both the naive PGD and the MSR3 algorithms
%(see Table \ref{table:comparison_of_algorithms}).

%We also develop a new meta-approach that can improve the performance of LME selection methods for any regularizer. Specifically, we extend the sparse relaxed regularized regression (SR3) framework~{(\cite{Zheng2019SR3})} to the LME setting. In linear regression, SR3 accelerates and improves the performance of  regularization strategies by introducing auxiliary variables that decouple the accuracy and sparsity requirements for the model coefficients.
%We develop a conceptual and algorithmic approach necessary to extend the SR3 concept to LME.  This development is necessary because the LME problem is nonlinear, nonconvex, and includes constraints on variance parameters.  We show that the new approach yields superior results in terms of specificity and sensitivity of feature selection, and is also computationally efficient.

All new methods are implemented in an open-source library called \texttt{pysr3}, which fills a gap for python mixed-models selection tools in \texttt{Python} (\cite{Buscemi2019Survey}, Table 3). Our algorithms are  1-2 orders of magnitude faster than available LASSO-based libraries for  mixed effects selection in \texttt{R},  see Table \ref{table:glmmlasso}. \texttt{pysr3} enables a standardized comparison of different methods in the LME setting, and makes both the PGD framework and its SR3 extension available to practitioners working with LME models. 
%\subsection{Notation}
%\begin{enumerate}
%\item sets like $\N$, $\B$, $\bS$, $\SS$.
%\item nonsmooth stuff, subdifferential and normal cones
%\end{enumerate}

%Roadmap
We begin in Section \ref{sec:LMEM} by giving a precise description of the LME
model and set the notation for the remainder of the paper. 
This is followed by a brief discussion of prior work on LMEs.
In Section \ref{sec:pgd_methods} we present our algorithms for the LME model 
starting with the naive PGD algorithm. This is followed by a description of the 
variable splitting technique used in \cite{Zheng2019SR3} to incorporate global 
variation information on the likelihood function into the direction finding subproblem
for the PGD algorithm. 
%This lays the foundation for the MSR3 algorithm. 
Next we tackle the problem of how to approximate the 
resulting optimal value function $u_\eta$ and its gradient where $\eta>0$ is the 
decoupling parameter. As noted, this is done using variable metric techniques and 
and interior point technology. We conclude Section \ref{sec:pgd_methods} 
with a discussion of the
MSR3 and MSR3-fast algorithms. In Section \ref{sec:applications} we discuss how
the underlying algorithmic parameters are set and test the algorithm on both 
simulated problems and a problem with real data. The paper is concluded in
Section \ref{sec:discussion} with a brief discussion of the contributions.

\edit{Add a paragraph summarizing the theory paper}
\edit{Add a paragraph summarizing software paper}

\section{Linear Mixed-Effects Models: Notation and Fundamentals}\label{sec:LMEM}
Mixed-effect models describe the relationship between an outcome variable and its predictors when the observations are grouped, for example in studies or clusters.  To set the notation, consider $m$ groups of observations indexed by $i$, with sizes $n_i$, and the total number of observations equal to $n = n_1 + n_2 + \dots + n_m$. For each group, we have design matrices for fixed features $X_i \in \R^{n_i \times p}$,  and matrices of random features $Z_i \in \R^{n_i \times q}$, along with vectors of outcomes $Y_i \in \R^{n_i}$. 
%Typically, columns of $Z_i$ are a subset of columns $X_i$ but it does not have to be the case in general. 
Let 
{$X = [X_1^T, X_2^T, \dots, X_m^T]^T$ and $Z = [Z_1^T, Z_2^T, \dots, Z_m^T]^T$.} 
Following~\cite{Patterson1971, Pinheiro2000}, we define a Linear Mixed-Effects (LME) model as
\begin{equation}
\begin{aligned}
	Y_i & = X_i\beta + Z_iu_i + \varepsilon_i, \quad i = 1 \dots m \\
	u_i & \sim \NN(0, \Gamma),\quad \Gamma \in \bS_{+}^{q} \\
	\varepsilon_i & \sim \NN(0, \Lambda_i), \quad \Lambda_i \in \bS_{++}^{n_i}
	\end{aligned}
	\label{eq:lme_setup}
\end{equation}
 where $\beta \in \R^p$ is a vector of fixed (mean) covariates, 
 $u_i \in \R^{q}$ are unobservable random effects assumed to be distributed normally with zero mean and the unknown covariance matrix $\Gamma$, and $\bS_{+}^{\nu}$ 
 and $\bS_{++}^{\nu}$ are the sets of
 real symmetric $\nu\times \nu$ positive semi-definite
 and positive definite matrices, respectively. 
Matrices $Z_i$ encode a wide variety of models, including
random intercepts ($Z_i$ are columns of 1's that add $u_i$ to all datapoints from the $i$th study)
and random slopes ($Z_i$ also scale $u_i$ according to the magnitude of a covariate), see e.g.~\cite{pinheiro2006mixed}. 
%In these typical examples, the columns of $Z_i$ are subsets of the columns
%of $X_i$.  
In our study, we assume that the observation error covariance matrices 
$\Lambda_i$ are given and that the random effects covariance matrix 
is an unknown diagonal matrix, i.e., $\Gamma = \Diag{\gamma}, \ \gamma \in \R^s_+$.
This assumption corresponds to the meta-analysis and meta-regression branch of mixed effects problems, 
which is the primary focus of our applied collaborations~ (see e.g. \cite{zheng2022burden,lescinsky2022health,razo2022effects,stanaway2022health,dai2022health}.)
The theoretical developments in this work allow extensions to other types of repeated measure models, 
but practical implementation requires significant additional effort, and we leave these extensions to future work.

% either
% assumed to have a simple parametric form, such as $\Lambda_i=\sig_i I$
% with $\sig_i$ to be determiined, or given given  
 %; we focus on the latter case below.

Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
recasts~(\ref{eq:lme_setup}) as a correlated noise model:
 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i.
}
For brevity, we refer to $\Omega_i(\Gamma)$ as just $\Omega_i$.
%without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
The reformulation \eqref{eq:lmm_correlated_noise_setup}
yields the following marginalized  negative log-likelihood function of a linear mixed-effects model~\citep{Patterson1971}:
\eq{
	\label{eq:lmm_objective}
	\LL_{ML}(\beta, \Gamma)  :=
%	 \textabove{$\Delta$}{=}
	 \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\ln{\det{\Omega_i}}.
}
Maximum likelihood estimates for $\beta$ and $\Gam$ are obtained by
solving the optimization problem
\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma)  \quad \mbox{s.t.} \quad \ \Gamma \in \bS_{+}^{q}.
}

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation

\begin{equation}
	\label{eq:beta_formula}
	\beta(\Gamma) = \argmin_{\beta}\LL(\beta, \Gamma) = \left(\sum_{i = 1}^m X_i^T\Omega_i^{-1}X_i\right)^{-1}\sum_{i = 1}^mX_i^T\Omega_i^{-1}y_i.
\end{equation}

By using the simplification $\Gam=\Diag{\gam}$, we obtain the problem
%{In the meta-analysis setting under study, 
%In our study, we assume that $\Gam$ is a diagonal matrix of the form
%$\Gamma = \Diag{\gamma}, \ \gamma \in \R^s$ with the 
%$\Lambda_i \in \R_{++}^{n_i \times n_i}$ fixed and given 
%%the positive definite constraint from (\ref{eq:ml_lme_optimization_setup}) transforms into a box constraint:
%so that our problem takes the form}
\eq{
	\label{eq:lme_diagonal_setup}
	\min_{\beta\in\R^p, \gam\in\R^q_+} & \ \LL(\beta,\gam)
	:=\LL_{ML}(\beta, \Diag{\gamma}) \, .
%	\\
%	s.t. & \ \gamma \geq 0\ .
	}
%Under the assumption $\Lambda_i \in \R_{++}^{n_i \times n_i}$ in~\eqref{eq:lme_setup}, 
In this setting, when an
entry $\gamma_j$ takes the  value $0$ the corresponding coordinates of all random effects $u_{ij}$ are identically $0$ for all $i$.

@@

This paper focuses the case where $\Gamma$ is diagonal, 
  (often referred to as \textit{the diagonal setup}) 
 and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely 
employed in epidemiological studies~\cite{murray2020global}. 
While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.

@@

Variable (feature) selection models seeks to select or rank the most important predictors in a dataset in order to get a parsimonious model at a minimal cost to prediction quality. 
Feature selection may be performed both on $\beta$, to find the sparse set of covariates that best explains the mean, and on $\gamma$, to find the sparse set of covariates that best accounts for 
variation between groups. Both types of selection have been studied in the literature, and both are accessible using the methods developed here.
%The selection process often occurs simultaneously with fitting the model, and a predictor is `selected' if its respective coefficient in the model is non-zero. 
If the desired number of coefficients $k$ is given, then the feature selection problem can be formulated as the minimization of a loss function $f(\theta)$ (e.g. the negative log-likelihood) subject to a zero-norm constraint:
\eq{
	\label{eq:general_feature_selection_zero_norm}
	&\min_{\theta}  f(\theta) \quad \mbox{ s.t. }  \quad \|\theta\|_0 \leq k 
	}
where $\|\theta\|_0$ denotes the number of nonzero entries in $\theta$, see panel (c) of Figure~\ref{fig:regularizers}.

@@

 \begin{figure}[h!]
    \centering
    \begin{tabular}{ccc}
        \includegraphics[width=0.3\textwidth]{figures/l1_regularizer.pdf}
&
           \includegraphics[width=0.3\textwidth]{figures/lh_regularizer.pdf}
    &
    \includegraphics[width=0.3\textwidth]{figures/l0_regularizer.pdf}\\
  %      \includegraphics[width=0.3\textwidth]{figs/scad_regularizer.pdf}\\
              (a) $\ell_1$  & (b) $\ell_{1/2}$ & (c) $\ell_0$
    \end{tabular}
    \caption{Common convex and non-convex regularizers used for feature selection.}
    \label{fig:regularizers}
\end{figure}

@@

 Feature selection for LMEs is more difficult than for linear regression models. 
 In linear regression the observations are independent, whereas in mixed-effects setup they are generally correlated. 
 In addition, LMEs have both mean effect variables  $\beta$ as well as random variance variables 
 $\Gamma$. % where the groups can have different relative importance. 
% Finally, selecting random features implies that $\Gamma$ will have zero columns and rows thereby forcing the optimization methods to the boundary of the feasible parameter set and making both the theoretical analysis and numerical computations more challenging.  
The shrinkage operator approach for linear regression 
\citep{Tibshirani1996} was first adapted 
to the problem of feature selection for the fixed effects in mixed-effect models by \cite{Lan2006}. 
The removal of a random effect from the model requires the elimination of an entire row and column from $\Gamma$. 
To make the problem more tractable, \cite{Chen2003} reparamtrized $\Gamma$ through a modified Cholesky decomposition $\Gamma(D,L): = DLL^TD$, where $D$ is a diagonal matrix and $L$ is a lower-triangular matrix with ones on the main diagonal, and focused on selecting elements of $D$. 
% In this case, one can select variance parameters by selecting the elements of the diagonal matrix $D$ while treating the non-diagonal entries of $L$ as free parameters. 
Based on this idea, \cite{Krishna2008} extended the Adaptive LASSO regularizer (\cite{Lan2006, Xu2015}) to mixed-effects setting using the objective
   \(
        \LL(\beta, \Gamma(D,L)) + \lambda\left(\sum_{i = 1}^p 
        \left|\frac{\beta_i}{\hat\beta_i}\right| + \sum_{j = 1}^q \frac{D_{ii}}{\hat{D_{ii}}} \right),
   \)
 where $\hat\beta$ and $\hat{D}$ are the solution of a non-penalized maximum likelihood problem and $\lambda$ is a tuning parameter for the weighted regularizer
 and is called the regularization parameter. 
 %This strategy has a well known issue: non-penalized estimates $\hat\beta$ and $\hat{D}$ may be inaccessible because the numerical algorithm may fail to converge, especially when the underlying solution is sparse. 
 \cite{Ibrahim2011} use a similar approach, penalizing  non-zero elements $\Gamma_{ij}$ directly. 
 Other methods that use Adaptive LASSO for simultaneous selection 
 of fixed and random effects are \cite{Lin2013, fan2014robust, pan2018simultaneous}. Adaptive LASSO is available to practitioners via \texttt{R} packages \texttt{glmmLasso}\footnote{\href{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}{https://rdrr.io/cran/glmmLasso/man/glmmLasso.html}} (\cite{groll2014variable}) and \texttt{lmmLasso}\footnote{\href{https://rdrr.io/cran/lmmlasso/}{https://rdrr.io/cran/lmmlasso/}}(\cite{schelldorfer2011estimation}).
